A Framework for Substructural Type Systems \thanks{Supported by organization x.}

### 2021-10-12 tex/esop22/examples.tex
[ writing ] add example term
[ new ] WRel example first draft

\subsection{A usage elaborator}

>>>>

To abbreviate this section, we use a simplified syntax compared to \name{}.
We allow for an arbitrary family of base types \AgdaBound{BaseTy}, and a single
type former \mbox{\ExecuteMetaData[\WReltex]{rAToB}}, equivalent to
\mbox{\ExecuteMetaData[\PaperExamplestex]{BangrAToB}} from the earlier system.

\ExecuteMetaData[\WReltex]{Ty}

In the term syntax, $\lambda$-abstraction now binds a variable with annotation
\AgdaBound{r}, while application needs to scale its argument by \AgdaBound{r}
(both in accordance with the function type they are acting on).

\ExecuteMetaData[\WReltex]{AnnArr}

In this subsection, we take the usage annotations to be the 4-element variance
posemiring.
We establish the property that all terms are monotonic in their free variables.
This monotonicity can be covariant or contravariant (or neither or both)
depending on the annotation of each free variable.
This provides an additional example to those of \citet{AbelBernardy2020}.
\todo{Cite before here.}

We will take semantics of this system into
\emph{worldly relations}~\cite{AbelBernardy2020}.
A worldly relation over a poset of worlds \AgdaBound{W} is a set over which
we have a \AgdaBound{W}-indexed binary relation satisfying a presheaf-like
property with respect to the order on \AgdaBound{W}.

\ExecuteMetaData[\WReltex]{WRel}

\begin{example}
  When \AgdaBound{W} is the 1-element set, a worldly relation is just a set
  equipped with a binary relation.
\end{example}

Morphisms between worldly relations \AgdaBound{R} and \AgdaBound{S} consist of
a mapping between the underlying sets such that that mapping preserves
relatedness from \AgdaBound{R} to \AgdaBound{S}.

\ExecuteMetaData[\WReltex]{WRelMor}

When the poset of worlds forms a (relational) commutative monoid, such worldly
relations support a symmetric monoidal closed structure.
We reuse the bunched connectives \AgdaRecord{$I^*$}, \AgdaRecord{$\sep$}, and
\AgdaRecord{$\wand$}, now over worlds rather than contexts.

\ExecuteMetaData[\WReltex]{IR}
\ExecuteMetaData[\WReltex]{tensorR}
\ExecuteMetaData[\WReltex]{lollyR}

The final piece of sematics we need is a \emph{bang} operator.
Instead of requiring extra algebraic structure on the worlds, we allow the
semantic \emph{bang} to be an arbitrary annotation-indexed functor on worldly
relations.
This functor must respect all of the structure on the indices, making it a
graded comonad over multiplication, as well as being lax monoidal at any
particular index \AgdaBound{r}.

\ExecuteMetaData[\WReltex]{Bang}

\begin{example}
  With \AgdaBound{W} being the 1-element set and annotations coming from the
  variance semiring, we can define the following \emph{bang}.
  It is always the identity on the set component, while the relation component
  consists of flipping the relation for contravariance and taking conjunctions
  to achieve both covariance and contravariance.
  When we want neither covariance nor contravariance, we use the always true
  predicate on worlds \AgdaFunction{$\dot1$}.

  \ExecuteMetaData[\Monotonicitytex]{BangR}
\end{example}

To associate semantics to syntax, we start as standard by associating worldly
relations to types.
We also extend the semantics of types to contexts, using \AgdaFunction{I$^R$},
\AgdaFunction{$\otimes^R$}, and \AgdaField{!$^R$} to interpret the empty
context, concatenation, and usage annotations on singletons, respectively.

\ExecuteMetaData[\WReltex]{sem}

The semantics of a term is then to be a morphism from the interpretation of the
context to the interpretation of the term's type.

\ExecuteMetaData[\WReltex]{sem-vdash}

Variables are given semantics by \AgdaFunction{lookup$^R$} (definition omitted).

\ExecuteMetaData[\WReltex]{lookupR-type}

Now, we give a \AgdaRecord{Semantics}.
The choice of \AgdaBound{$\V$} as
\AgdaRecord{\AgdaUnderscore{}$\sqni$\AgdaUnderscore{}} is somewhat arbitrary,
given that a standard denotational semantics would not use intermediate
environments in the same sense as renaming and substitution, but allows us to
reuse the standard facts that variables support renaming and identity
environments.
With this choice of \AgdaBound{$\V$} and \AgdaBound{$\C$}, we interpret
environment entries by \AgdaFunction{lookup$^R$}.
Meanwhile, for the logical rules, we ignore environments by using
\AgdaFunction{reify} to just deal with morphisms in an extended context.
As such, $\lambda$-abstractions are easy to interpret, while applications
require some massaging to remove the extension by an empty context, followed by
some plumbing to split the interpretation of the context according to the usage
constraints and feed the interpretation of the argument \AgdaBound{n$'$} into
the interpretation of the function \AgdaBound{m$'$}.

\ExecuteMetaData[\WReltex]{Wrel}

In order to map open terms to interpretations, we take the action of the
semantics and give the identity renaming as the starting environment.

\ExecuteMetaData[\WReltex]{wrel}

\begin{example}
  We can make a subtraction function from primitive addition and negation on
  integers.
  Subtraction is covariant in its first argument and contravariant in its
  second argument.
  We give the definition in pseudocode, as we have not yet seen how to
  conveniently write terms (\cref{sec:usage-elaborator}).

  \begin{align*}
    &{\sim\sim}p :
      {\uparrow\uparrow}\mathbb Z \multimap
      {\uparrow\uparrow}\mathbb Z \multimap \mathbb Z,
      {\sim\sim}n : {\downarrow\downarrow}\mathbb Z \multimap \mathbb Z
      \vdash \mathnormal{minus} :
      {\uparrow\uparrow}\mathbb Z \multimap
      {\downarrow\downarrow}\mathbb Z \multimap
      \mathbb Z
    \\
    &\mathnormal{minus} \coloneqq \lambda x.~\lambda y.~p\,x\,(n\,y)
  \end{align*}

  We observe that the set component of this term's semantics is just the
  expected Agda function when the two free variables are given appropriate
  interpretations.

  \ExecuteMetaData[\Monotonicitytex]{minus-set}

  Furthermore, the relational component of the semantics yields the free
  theorem that the Agda subtraction so defined is monotonic in the expected way.
  This relies on library proofs that addition and negation are suitably
  monotonic.

  \ExecuteMetaData[\Monotonicitytex]{thm}
\end{example}

\subsection{A usage elaborator}\label{sec:usage-elaborator}

### 2021-10-12 tex/esop22/quant-framework.tex
[ writing ] explaining semantics
[ refactor ] example term in separate file
[ new ] WRel example first draft

\def\genericlr{../../generic-lr/src/processed-latex}
\def\Syntaxtex{\genericlr/Generic/Linear/Syntax.tex}
\def\Interpretationtex{\genericlr/Generic/Linear/Syntax/Interpretation.tex}
\def\Maptex{\genericlr/Generic/Linear/Syntax/Interpretation/Map.tex}
\def\Termtex{\genericlr/Generic/Linear/Syntax/Term.tex}
\def\Semanticstex{\genericlr/Generic/Linear/Semantics.tex}
\def\Syntactictex{\genericlr/Generic/Linear/Semantics/Syntactic.tex}
\def\Renamingtex{\genericlr/Generic/Linear/Renaming.tex}
\def\PaperExamplestex{\genericlr/Generic/Linear/Example/PaperExamples.tex}
\def\UsageChecktex{\genericlr/Generic/Linear/Example/UsageCheck.tex}
\def\Snippetstex{../../agda/processed-latex/Snippets.tex}

@@

\keywords{First keyword  \and Second keyword \and Another keyword.}
\end{abstract}
%
%
%
\begin{enumerate}
\item (Section 1)
  \begin{enumerate}
  \item   Motivate intrinsically typed syntax (in Agda)
    \begin{enumerate}
    \item Show the ``GADT'' for well scoped syntax
    \item Show the well typed version
    \end{enumerate}
  \item Motivate adding posemiring annotations
    \begin{enumerate}
    \item Pointing to the existing work (Brunel et al., Ghica and Smith, Granule, previous of ours, Bernardy and Abel)
    \item The fact that we can encode lots of other systems (DILL, PD, monotonicity, information flow, differential privacy)
    \end{enumerate}
  \end{enumerate}
\item (Section 2) Background / notation fixing for:
  \begin{enumerate}
  \item Partially ordered semirings
  \item Vectors
  \item Linear maps and linear relations (as a generalisation of $P \leq Q\Psi$ which is easier to work with)
  \end{enumerate}
\item (Section 3) Bunched Typing Rules
  \begin{enumerate}
  \item Introduces $\lambda\mathcal{R}$
  \item Rephrases the typing rules in terms of bunches (as well as implicit contexts)
  \end{enumerate}
\item (Section 4) Universe of Generic Usage-annotated syntax
  \begin{enumerate}
  \item Premises
  \item Rules
  \item Systems
  \item Example(s)
  \item Intepretation as Terms
  \item Examples of Terms; using pattern synonyms (point out that proofs of usage accounting are painful)
  \end{enumerate}
\item (Section 5) Example Syntaxes
  \begin{enumerate}
  \item $\lambda\mathcal{R}$ (incl DILL, PD, differential privacy, monotonicity)
  \item Inductive types with a well resourced fold
  \item Bernardy-Abel style
  \item First-order (circuitry)
  \item LNL
  \item $\lambda\mu\tilde{\mu}$
  \end{enumerate}
\item (Section 6) Environments
  \begin{enumerate}
  \item Motivate in terms of renaming and substitution for simple types (where it is just a function)
  \item Motivate the general idea of using linear relations (i.e. transformations + weakening)
    \begin{enumerate}
    \item Key idea / slogan : environment = change of basis
    \end{enumerate}
  \item Definition of environments
  \item Examples
  \item Properties, incl. preservation of linear operations
  \end{enumerate}
\item (Section 7) Semantics, or How to apply an Environment to a Term
  \begin{enumerate}
  \item Renameability
  \item $\mathsf{semantics}$
  \end{enumerate}
\item (Section 8) Examples of Semantics
  \begin{enumerate}
  \item Renaming and Substitution (term-valued)
  \item Denotational Semantics (relational world-indexed (\texttt{Generic/Linear/Example/WRel.agda})
  \item Usage Elaborator
  \end{enumerate}
\item (Section 9) Related Work
\item (Section 10) Conclusion
\end{enumerate}

>>>>

\def\genericlr{../../generic-lr/src/processed-latex}
\def\Syntaxtex{\genericlr/Generic/Linear/Syntax.tex}
\def\Interpretationtex{\genericlr/Generic/Linear/Syntax/Interpretation.tex}
\def\Maptex{\genericlr/Generic/Linear/Syntax/Interpretation/Map.tex}
\def\Termtex{\genericlr/Generic/Linear/Syntax/Term.tex}
\def\Semanticstex{\genericlr/Generic/Linear/Semantics.tex}
\def\Syntactictex{\genericlr/Generic/Linear/Semantics/Syntactic.tex}
\def\Renamingtex{\genericlr/Generic/Linear/Renaming.tex}
\def\PaperExamplestex{\genericlr/Generic/Linear/Example/PaperExamples.tex}
\def\HeavyItex{\genericlr/Generic/Linear/Example/PaperExamples/HeavyI.tex}
\def\UsageChecktex{\genericlr/Generic/Linear/Example/UsageCheck.tex}
\def\WReltex{\genericlr/Generic/Linear/Example/WRel.tex}
\def\Monotonicitytex{\genericlr/Generic/Linear/Example/WRel/Monotonicity.tex}
\def\Snippetstex{../../agda/processed-latex/Snippets.tex}

@@

\keywords{First keyword  \and Second keyword \and Another keyword.}
\end{abstract}
%
%
%
\begin{enumerate}
\item (Section 1)
  \begin{enumerate}
  \item   Motivate intrinsically typed syntax (in Agda)
    \begin{enumerate}
    \item Show the ``GADT'' for well scoped syntax
    \item Show the well typed version
    \end{enumerate}
  \item Motivate adding posemiring annotations
    \begin{enumerate}
    \item Pointing to the existing work (Brunel et al., Ghica and Smith, Granule, previous of ours, Bernardy and Abel)
    \item The fact that we can encode lots of other systems (DILL, PD, monotonicity, information flow, differential privacy)
    \end{enumerate}
  \end{enumerate}
\item (Section 2) Background / notation fixing for:
  \begin{enumerate}
  \item Partially ordered semirings
  \item Vectors
  \item Linear maps and linear relations (as a generalisation of $P \leq Q\Psi$ which is easier to work with)
  \end{enumerate}
\item (Section 3) Bunched Typing Rules
  \begin{enumerate}
  \item Introduces $\lambda\mathcal{R}$
  \item Rephrases the typing rules in terms of bunches (as well as implicit contexts)
  \end{enumerate}
\item (Section 4) Universe of Generic Usage-annotated syntax
  \begin{enumerate}
  \item Premises
  \item Rules
  \item Systems
  \item Example(s)
  \item Intepretation as Terms
  \item Examples of Terms; using pattern synonyms (point out that proofs of usage accounting are painful)
  \end{enumerate}
\item (Section 5) Example Syntaxes
  \begin{enumerate}
  \item $\lambda\mathcal{R}$ (incl DILL, PD, differential privacy, monotonicity)
  \item Inductive types with a well resourced fold
  \item Bernardy-Abel style
  \item First-order (circuitry)
  \item LNL
  \item $\lambda\mu\tilde{\mu}$
  \end{enumerate}
\item (Section 6) Environments
  \begin{enumerate}
  \item Motivate in terms of renaming and substitution for simple types (where it is just a function)
  \item Motivate the general idea of using linear relations (i.e. transformations + weakening)
    \begin{enumerate}
    \item Key idea / slogan : environment = change of basis
    \end{enumerate}
  \item Definition of environments
  \item Examples
  \item Properties, incl. preservation of linear operations
  \end{enumerate}
\item (Section 7) Semantics, or How to apply an Environment to a Term
  \begin{enumerate}
  \item Renameability
  \item $\mathsf{semantics}$
  \end{enumerate}
\item (Section 8) Examples of Semantics
  \begin{enumerate}
  \item Renaming and Substitution (term-valued)
  \item Denotational Semantics (relational world-indexed (\texttt{Generic/Linear/Example/WRel.agda}))
  \item Usage Elaborator
  \end{enumerate}
\item (Section 9) Related Work
\item (Section 10) Conclusion
\end{enumerate}
